---
title: 'Descriptive Statistics'
subtitle: 'Univariate Statistics'
author: "Karol Flisikowski"
date: "`r Sys.Date()`"
output:
  rmdformats::downcute:
    self_contained: true
    thumbnails: false
    lightbox: true
    gallery: true
    highlight: pygments
---

```{r setup1, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(qwraps2_markup = "markdown")
library(qwraps2)
library(arsenal)
library(e1071)
library(haven)
library(papeR)
library(dplyr)
library(tidyverse)
library(ggplot2)
library(kableExtra)
library(summarytools)
library(classInt)
library(pastecs)
library(reporttools)
library(desctable)
library(frequency)
library(ggpubr)
download.file("https://github.com/kflisikowski/ds/blob/master/data_apartments.csv?raw=true", destfile ="mieszkania.csv",mode="wb")
mieszkania <- read.csv("mieszkania.csv",sep=";",dec=",")
```

## Data

In our example this week, we are going to use the fake data - about real estates in Wroclaw - prices by districts, size of apartments and many more.

## Data wrangling

As you can see, not all formats of our variables are adapted. We need to prepare appropriate formats of our variables according to their measurement scale and future application.

```{r wrangling, include=TRUE}
mieszkania$district<-as.factor(mieszkania$district)
mieszkania$building_type<-as.factor(mieszkania$building_type)
mieszkania$rooms<-factor(mieszkania$rooms,ordered=TRUE)
attach(mieszkania)
mieszkania$price_PLN<-as.numeric(mieszkania$price_PLN)
mieszkania$price_EUR<-as.numeric(mieszkania$price_EUR)
```

## Frequency Tables and TAI

In the first step of our analysis, we will group our data into a simple frequency table.

First, let's look at the distribution of housing prices in our sample and verify tabular validity using the TAI measure:

```{r table, message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
etykiety<-c("350-450 kPLN","450-550 kPLN","550-650 kPLN","650-750 kPLN","750-850 kPLN","850-950 kPLN","950-1050 kPLN","1050-1150 kPLN","1150-1250 kPLN","1250-1350 kPLN")
limits<-cut(mieszkania$price_PLN,seq(350000,1350000,by=100000),labels=etykiety)
tabela1<-freq(limits,type="html")
```

Ok, it looks quite ugly, so let's wrap it up using the 'kable' package:

```{r tai, echo=FALSE}
kbl(tabela1,caption = "Apartments in Wroclaw - prices in kPLN") %>%
    kable_material(c("striped", "hover"))
tab1<-classIntervals(mieszkania$price_PLN,n=10,style="fixed",fixedBreaks=seq(350000,1350000,by=100000))
jenks.tests(tab1)
```

As we can see - the TAI index is quite high. 0.85 means that we can accept the proposed construction of the frequency table.

## Basic plots

In this section, we should represent our data using basic (pre-installed in R) graphics. Select the most appropriate graphs depending on the scale of the selected variables. Explore the heterogeneity of the distribution by presenting the data by group (e.g., by neighborhood, building type, etc.). Don't forget about main titles, labels and legends. Read more about graphical parameters [here](http://www.sthda.com/english/wiki/graphical-parameters).

```{r histogram, echo=FALSE}
hist(price_PLN, breaks="FD", col="green", probability = TRUE,
     main="Ceny w PLN - Wroclaw")
lines(density(price_PLN[district=="Krzyki"]),col=2)
lines(density(price_PLN[district=="Biskupin"]),col=3)
lines(density(price_PLN[district=="Srodmiescie"]),col=4)
legend("topright", legend=c("Krzyki", "Biskupin", "Srodmiescie"),
       col=c(2,3,4), lty=1:2, horiz=FALSE, box.lty=0, cex=0.8)

```

Note that the `echo = FALSE` parameter has been added to the code snippet to prevent printing the R code that generated the graph.

```{r boxplot, echo=FALSE}
boxplot(price_PLN~district)
```

## ggplot2 plots

Now, let's use the ***ggplot2*** and ***ggpubr*** libraries to plot.

```{r histogram2, echo=FALSE}
# Density plot of "price_PLN"
#::::::::::::::::::::::::::::::::::::::
density.p <- ggdensity(mieszkania, x = "price_PLN", 
                       fill = "district", palette = "jco")+
  stat_overlay_normal_density(color = "red", linetype = "dashed")

# Draw the summary table of price_PLN
#::::::::::::::::::::::::::::::::::::::
# Compute descriptive statistics by groups
stable <- desc_statby(mieszkania, measure.var = "price_PLN",
                      grps = "district")
stable <- stable[, c("district", "length", "mean", "sd")]
# Summary table plot, medium orange theme
stable.p <- ggtexttable(stable, rows = NULL, 
                        theme = ttheme("mOrange"))
# Draw text
#::::::::::::::::::::::::::::::::::::::
text <- paste("Price per apartment by 3 districts - Wroclaw.",
              "Random sample of 200 apartments.",
               sep = " ")
text.p <- ggparagraph(text = text, face = "italic", size = 11, color = "black")
# Arrange the plots on the same page
ggarrange(density.p, stable.p, text.p, 
          ncol = 1, nrow = 3,
          heights = c(1, 0.5, 0.3))
```

Ggplot2 allows you to show the average value for each group using the **stat_summary()** function. You no longer need to calculate average values before creating a graph!

```{r boxplot2, echo=FALSE}
ggplot(mieszkania, aes(x=district, y=price_PLN)) +
    geom_boxplot(alpha=0.7) +
    stat_summary(fun="mean", geom="point", shape=20, size=5, color="red", fill="red") +
 geom_jitter() +
    facet_grid(~building_type) +
    scale_fill_brewer(palette="Set1")

```

## Faceting

Faceting generates small multiples, each showing a different subset of the data. They are a powerful tool for exploratory data analysis: you can quickly compare patterns in different parts of the data and see if they are the same or different. Read more [here](https://ggplot2-book.org/facet.html).

```{r facet1, echo=FALSE}
plot1 <- ggplot(mieszkania, aes(price_PLN, rooms)) + 
  geom_abline() +
  geom_jitter(width = 0.1, height = 0.1) 
plot1 + facet_wrap(~district)
```

## Univariate Statistics

Before automatically reporting the full summary table of descriptive statistics, this time your goal is to measure the central tendency of the price distribution. Compare the mean, median, and mode along with positional measures - quantiles - by neighborhood and building type or number of rooms in an apartment.

```{r}
    mean(price_PLN)
    median(price_PLN)
    sd(price_PLN) #standard deviation
    var(price_PLN) #variance
    coeff_var<-sd(price_PLN)/mean(price_PLN) #coefficient of variability %
    coeff_var
    IQR(price_PLN)# difference between quartiles =Q3-Q1 
    sx<-IQR(price_PLN)/2  #interquartile deviation
    coeff_varx<-sx/median(price_PLN) #IQR coefficient of variability %
    coeff_varx
    min(price_PLN)
    max(price_PLN)
    quantile(price_PLN,probs=c(0,0.1,0.25,0.5,0.75,0.95,1),na.rm=TRUE)
```


Ok, we have calculated all of the basic summary statistics above. Let's wrap them up together now.


```{r kable_report, echo=FALSE}
mieszkania_list <- split(mieszkania$price_PLN, mieszkania$rooms)
inline_plot <- data.frame(rooms = c(1, 2, 3, 4), boxplot = "", histogram = "",
                          line1 = "", line2 = "", points1 = "")
inline_plot %>%
  kbl(booktabs = TRUE) %>%
  kable_paper(full_width = FALSE) %>%
  column_spec(2, image = spec_boxplot(mieszkania_list)) %>%
  column_spec(3, image = spec_hist(mieszkania_list)) %>%
  column_spec(4, image = spec_plot(mieszkania_list, same_lim = TRUE)) %>%
  column_spec(5, image = spec_plot(mieszkania_list, same_lim = FALSE)) %>%
  column_spec(6, image = spec_plot(mieszkania_list, type = "p"))

```

Ok, now we will finally summarize the basic measures of central tendency for prices by neighborhood/building type using the '***kable***' package. Feel free to customize your final report. See some hints [here](https://cran.r-project.org/web/packages/qwraps2/vignettes/summary-statistics.html).

```{r kable_report2, echo=FALSE, message=FALSE, warning=FALSE}
library(psych)
raport <-
  list("Price in PLN" =
       list("Min"       = ~ min(price_PLN),
            "Max"       = ~ max(price_PLN),
            "Q1"        = ~ quantile(price_PLN,0.25),
            "Median" = ~ round(median(price_PLN),2),
            "Q3"        = ~ quantile(price_PLN,0.75),
            "Mean" = ~ round(mean(price_PLN),2),
            "Sd" = ~ round(sd(price_PLN),2),
             "IQR" = ~ round(iqr(price_PLN),2),
            "Sx" = ~ round(iqr(price_PLN)/2,2),
            "Var %" = ~ round((sd(price_PLN)/mean(price_PLN)),2),
            "IQR Var %" = ~ round((iqr(price_PLN)/median(price_PLN)),2),
            "Skewness" = ~  round(skew(price_PLN),2),
             "Kurtosis" = ~  round(kurtosi(price_PLN),2)
            ))
tabela<-summary_table(mieszkania, summaries = raport, by = c("rooms"))

kbl(tabela,
  digits = 2,
  caption="Table 1. Apartments in Wroclaw - prices in PLN by number of rooms.",
  col.names = c('1 room', '2 rooms', '3 rooms', '4 rooms'))%>%
 kable_classic(full_width = F, html_font = "Cambria")%>%
 kable_styling(bootstrap_options = c("striped", "hover"))
```

Your task this week is to interpret all of the basic measures we have calculated above. Feel free to change this report: adjust formats, add plots and charts etc. 

We are going to discuss the results next week together during our class. See you soon! 

